Qualitative Probabilistic Programming

### 2015-02-25 paper/paper.tex
paper

\documentclass[proceed]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\setlength{\emergencystretch}{10em}

@@

  \section{Introduction}

@@

  In the following, we first specify the syntax used to write Quipp programs,
  including the notation for unknown variables and functions.
  We describe the class of exponential family variables and functions that our system can learn,
  and present the expectation maximization algorithm used to learn them.
  We then demonstrate the expressiveness of our language, and the broad
  applicability of our algorithm, by writing some of the most common machine learning models
  in Quipp: clustering, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
  a neural net.

  \section{Syntax}

  Quipp is implemented as a Python library, so Quipp programs are written as Python programs that have access
  to special functions.  Here is an example of a Quipp program to perform clustering:

@@

  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{get\_point}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the disttribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{get\_point} function with a learned unknown linear function with noise.

  \section{Class of Distributions}

    For unknown functions, the class of random functions used is a kind
    of generalized linear model.  We assume that the result
    is distributed from some exponential family whose natural
    parameters are determined from the arguments.  We label
    some subset of the sufficient statistics as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector x of the
    features of $x$.  Then

    $$p_{\mathbf{N}}(y | x) = \exp(\psi(x)^T N \phi(y) - g(N^T \psi(X)))$$

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem:

    \begin{align*}
      \log p_{\mathbf{N}}(y | x) = \psi(x)^T N \phi(y) - g(N^T \psi(X))
    \end{align*}
    since the log probability function is concave when $g$ is convex (as it is
    for any exponential family distribution).

    Therefore, it is possible to use Newton's method to optimize the parameters.

  \section{Inference}

    To perform inference of both latent variables and parameters, we first
    translate the program to a factor graph.  Next, we run the
    expectation maximization algorithm on this graph, iterating stages of
    estimating latent variables using Metropolis Hastings and parameter
    inference using Newton's method.

    In the factor graph, there are 2 types of variables: real numbers
    and categorical values.
    In Quipp, we have an additonal tuple type, which gets translated
    to multiple underlying variables in the factor graph.

    There are a few different types of factors:
    \begin{enumerate}
      \item Factors representing known distributions
      \item Factors representing unknown distribution
      \item Factors asserting that a value is equal to some constant
    \end{enumerate}

    How is inference in the factor graph performed?  We use the Metropolis Hastings
    algorithm, updating one node at a time.  To get a proposal for a given node,
    we approximate the local distribution using message passing (similar to
    in variational message passing).  Specifically, for each factor
    $f(x_1, x_2, ..., x_n)$ and each $1 \leq i \leq n$, we send a message
    to variable $x_i$ that approximates $x_i \mapsto f(x_1, x_2, ..., x_n)$.

  \section{Examples}

  \subsection{Clustering}
\begin{verbatim}
dim = 2
nclusters = 3

@@

def sample():
  cluster = get_cluster()
  return (cluster, get_point(cluster))
\end{verbatim}

@@

\begin{center}
\includegraphics[scale=0.5]{cluster_accuracy.png}
\end{center}

\subsection{Naive Bayes}

\begin{verbatim}
nfeatures = 10
FeaturesType = Vector(nfeatures, Bool)

@@

def sample():
  which_class = get_class()
  if which_class:
    features = class_1_features()
  else:
    features = class_2_features()
  return (which_class, features)
\end{verbatim}

@@

  \subsection{Factor Analysis}

\begin{verbatim}
num_components = 3
point_dim = 5

@@

def sample():
  components = [normal(0, 1) for i in range(num_components)]
  return (comenents, get_point(components))
\end{verbatim}

@@

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{Hidden Markov model}
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

@@

def sample():
  states = [uniform(num_states)]
  for i in range(sequence_length - 1):
    states.append(trans_fun(states[-1]))
  return (states, [obs_fun(s) for s in states])
\end{verbatim}

@@

\subsection{Latent Dirichlet Allocation}

\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

@@

def sample():
    which_class = uniform_categorical(n_classes)
    words = [class_to_word(which_class) for i in range(n_words_per_document)]
    return (which_class, words)
\end{verbatim}

@@

\subsection{Neural Network}

\begin{verbatim}
input_dim = 100
hidden_dim = 20
output_dim = 2

@@

def sample(input_layer):
  hidden_layer = input_to_hidden(input_layer)
  output_layer = hidden_to_output(hidden_layer)
  return ((), output_layer)
\end{verbatim}

\subsection{A more complex model}

TODO: come up with a more complex model.  Show how this model can be iteratively improved by adding information.  It would make sense for this to be some sort of data science type

  \section{Discussion}

  - Comments about results
  - Future directions
    - Unbounded models
    - Recursive algebraic data types

>>>>

\documentclass[proceed]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\setlength{\emergencystretch}{10em}
\usepackage{proceed2}

@@

  \section{INTRODUCTION}

@@

  In the following, we first specify the syntax used to write Quipp programs,
  including the notation for unknown variables and functions.
  We describe the class of exponential family variables and functions that our system can learn,
  and present the expectation maximization algorithm used to learn them.
  We then demonstrate the expressiveness of our language, and the broad
  applicability of our algorithm, by writing some of the most common machine learning models
  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
  a neural net.

  \section{SYNTAX}

  Quipp is implemented as a Python library, so Quipp programs are written as Python programs that have access
  to special functions.  Here is an example of a Quipp program to cluster 2d points into 3 clusters:

@@

  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{get\_point}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the distribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{get\_point} function with a learned function that will take a random
  sample from the given cluster.

  \section{FAMILY OF DISTRIBUTIONS}

  For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the result
    is distributed from some exponential family whose natural
    parameters are determined from the arguments:

    $$p_{\eta}(y | x) = \exp\left(\eta(x)^T \phi(y) - g(\eta(x))\right)$$

    Here, $\eta$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
    and $g$ is the log partition function.

    To determine $\eta$ as a function of $x$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.  Features are the same as sufficient
    statistics for the categorical distribution, but while both $X$ and $X^2$ are
    sufficient statistics for the Gaussian distribution, only $X$ is a feature.

    Let $\psi(x)$ be the features of $x$.  Then
    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for entries not in the first
    row corresponding to sufficient statistics of $y$ that are not features.  This ensures that only the
    natural parameters for features are affected by $x$.

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem because the log probability function is concave:
    $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$

    This relies on the fact that $g$ is convex, but this is true in general for any exponential family distribution.
    Since the problem is convex, it is possible to use Newton's method to optimize the parameters.  Although
    the only exponential family distributions we use in this paper are the categorical and Gaussian distributions,
    we can use the same algorithms for other exponential families, such as the Poisson and gamma distributions.

  \section{INFERENCE}

    To infer both latent variables and parameters, we first
    translate the program to a factor graph.  Next, we run the
    expectation maximization algorithm on this graph, iterating stages of
    estimating latent variables using Metropolis Hastings and inferring
    parameters using Newton's method.

    In the factor graph, there are 2 types of variables: real numbers
    and categorical values.
    In Quipp, we have an additonal tuple type, which gets translated
    to multiple underlying variables in the factor graph.  In the
    clustering example, we would represent each point by 2 real number variables,
    and each cluster assignment as a single categorical variable (with 3 alternatives).

    Factors can represent known distributions,
    unknown distributions, or observations.
    In the clustering example, we will have, for each cluster assignment variable,
    a factor representing the fact that it comes from a (known) uniform distribution.
    For each real number in a data point, we will have a factor connecting it to
    the cluster assignment for the point, which will represent the unknown distribution.
    This factor depends on the specific values of the parameters for the $\texttt{get\_point}$ function.
    Also, since data points are observed, they will have constant factors attached to them
    representing these observations.

    \begin{figure}[h]
    \begin{center}
    \includegraphics[scale=0.2]{cluster_diagram.png}
    \end{center}
    \caption{Translating clustering model to factor graph.  TODO fix variable names} 
    \end{figure}

    How is inference in the factor graph performed?  We use the Metropolis Hastings
    algorithm, updating one node at a time.  To get a proposal for a given node,
    we approximate the local distribution using message passing (similar to
    in variational message passing).  Specifically, for each factor
    $f(x_1, x_2, ..., x_n)$ and each $1 \leq i \leq n$, we send a message
    to variable $x_i$ that approximates $x_i \mapsto \log f(x_1, x_2, ..., x_n)$ as
    an affine function as $\phi(x_i)$.  The proposal distribution is then the
    product of these messages, which will approximate the local distribution of
    the variable.

  \section{EXAMPLES}

  \subsection{CLUSTERING}
{\small
\begin{verbatim}
dim = 2
nclusters = 3

@@

def sample():
  cluster = get_cluster()
  return (cluster, get_point(cluster))
\end{verbatim}
}

@@

\begin{center}
\includegraphics[scale=0.4]{cluster_accuracy.png}
\end{center}

\subsection{NAIVE BAYES}

{\small
\begin{verbatim}
nfeatures = 10
FeaturesType = Vector(nfeatures, Bool)

@@

def sample():
  which_class = get_class()
  if which_class:
    features = class_1_features()
  else:
    features = class_2_features()
  return (which_class, features)
\end{verbatim}
}

@@

  \subsection{FACTOR ANALYSIS}
{\small
\begin{verbatim}
num_components = 3
point_dim = 5

@@

def sample():
  components = [normal(0, 1)
                for i in range(num_components)]
  return (comenents, get_point(components))
\end{verbatim}
}

@@

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{HIDDEN MARKOV MODEL}
{\small
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

@@

def sample():
  states = [uniform(num_states)]
  for i in range(sequence_length - 1):
    states.append(trans_fun(states[-1]))
  return (states, [obs_fun(s) for s in states])
\end{verbatim}
}

@@

\subsection{LATENT DIRICHLET ALLOCATION}
{\small
\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

@@

def sample():
    which_class = uniform_categorical(n_classes)
    words = [class_to_word(which_class)
             for i in range(n_words_per_document)]
    return (which_class, words)
\end{verbatim}
}

@@

\subsection{NEURAL NETWORK}

{\small
\begin{verbatim}
input_dim = 100
hidden_dim = 20
output_dim = 2

@@

def sample(input_layer):
  hidden_layer = input_to_hidden(input_layer)
  output_layer = hidden_to_output(hidden_layer)
  return ((), output_layer)
\end{verbatim}
}

\subsection{A MORE COMPLEX MODEL}

TODO: if there is time, we should put a more complex data science type example, where we add dependencies and show change in accuracy as we add/remove assumptions.

  \section{DISCUSSION}

  We have found that it is possible to write many useful machine learning models as Quipp programs and then use generic algorithms for inference.  Furthermore, performance is <???>.  This should make it much easier for non-experts to write useful machine learning models.

  In the future, it will be useful to support unbounded recursion in models (as in Church).  Furthermore, it will also be useful to expand the set of types supported.  It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.
